Associations between Chlorophyll *a* and various Microcystin-LR Health Advisory

### 2015-10-05 manuscript/manuscript.Rmd

Over the last decade or so numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; @HABHRCA2014].  Recently, the US Environmental Protection Agency (USEPA) has released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic; @usepa2015drinking].  Additionally, the World Health Organization (WHO) has had proposed advisory levels for drinking water and a range of recreational risk levels [@world2003cyanobacterial; @chorus1999toxic].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

One of these complications is that they rely on available measurements of microcystin-LR.  While laboratory testing remains the gold standard for quantifying microcystin-LY concentrations in water samples, several field test kits have been developed.  Even though field tests provide a much needed means for rapid assessment, they are infrequently available to lake managers.  These kits are moderately expensive (approximately $150-$200 depending on specific kit) with a limited shelf life (typically one year) [@james2011environmental; @aranda2015evaluation].  Additionally, each technique requires nuanced understanding of the detection method (e.g., limit of detection, specific microcystin variants being measured, and sampling protocol).    Fortunately, microcystin-LR has been shown to be associated with several other, more commonly measured and understood components of water quality.

Chlorophyll *a* is a very commonly measured components of water quality that is also known to be associated with Microsystin-LR concentrations [@lee2000variation; @paerl2013harmful; heisler2008eutrophication].  Additionally there are many rapid measurements for assessing chlorophyll *a* levels *in situ*.  For instance, there are small or hand held flourometers that measure chlorohpyll *a*.   Given these facts, it might be possible to identify chlorophyll *a* concentrations that would be associated with the various Microcystin-LR health advisory levels.  Identifying these associations would provide another reliable tool for water resource managers to use to help manage the threat to public health posed by cHABs and would be especially useful in the absence of microcystin-LR concentrations.  Thus, the goal of this paper is to utlize the National Lakes Assessment data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [@usepa2009national].  So that others may repeat or adjust this analysis, the data, code, and this manuscript are freely available via [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

@@

A detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for details.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  The 95% confidence intervals were calculated from 1000 bootstrapped samples. To identify chlorophyll *a* concentrations of concern we used a 50% conditional probability of exceeding each health advisory level and extracted the minimum chlorophyll *a* concentration that was associated with the upper confidence level being 50% or greater.  As both microcystin-LR and chlorophyll *a* values were both highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)`.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)` and the mean was `r median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)`.  Lastly, of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA childrens drinking water standard, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA adult drinking water standard,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO drinking water standard,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO recreational standard. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)`.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` and the mean was `r mean(nla_dat$CHLA,na.rm=T)`.  The associations between chlorophyll *a* and the upper confidence interval with a conditional probability of 50% were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]` for 0.3, 1.0, 1.6 and 2.0 $\mu$g/L, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

@@

There are numerous possible uses for the chlorophyll *a* and microcystin-LR advisory cut-off values.  First, in the absence of microcystin-LR measurements, exceedence of the chlorophyll *a* concentrations could be a trigger for further actions.  Given that there is uncertainity around these chlorophyll *a* cutoffs the best case scenario would be to monitor for chlorophyll and in the event of exceeding a target concentration take water samples and have those samples tested in a lab for microcystin-LR.  A second potential use is to identify possible bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanism responsible for adverse impacts (e.g. toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs one approach has been to identify an increase over a baseline concentration of phycocyanin [@miller2013spatiotemporal].  This is a useful approach for targeted studies, but phycocyanin is also not always available and measures the predominance of cyanobacterial pigments and not toxins. Using our chlorophyll *a* cutoffs provides a value that is more directly associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.

>>>>

Over the last decade or so numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; @HABHRCA2014], and the US Environmental Protection Agency (USEPA) has recently released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic; @usepa2015drinking].  Additionally, the World Health Organization (WHO) has had proposed advisory levels for drinking water and a range of recreational risk levels [@world2003cyanobacterial; @chorus1999toxic].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

One of these complications is that they rely on available measurements of microcystin-LR.  While laboratory testing remains the gold standard for quantifying microcystin-LR concentrations in water samples, several field test kits have been developed.  Even though field tests provide a much needed means for rapid assessment, they are not yet widely used and are moderately expensive (approximately $150-$200 depending on specific kit) with a limited shelf life (typically one year) [@james2011environmental; @aranda2015evaluation].  Additionally, each technique requires nuanced understanding of the detection method (e.g., limit of detection, specific microcystin variants being measured, and sampling protocol).    Fortunately, microcystin-LR has been shown to be associated with several other, more commonly measured and understood components of water quality.

Chlorophyll *a* is a very commonly measured component of water quality that is also known to be associated with Microsystin-LR concentrations [@lee2000variation; @paerl2013harmful; heisler2008eutrophication].  Additionally there are many rapid measurements for assessing chlorophyll *a* levels *in situ*.  For instance, there are small or hand held flourometers that measure chlorohpyll *a*.   Given these facts, it might be possible to identify chlorophyll *a* concentrations that would be associated with the various Microcystin-LR health advisory levels.  Identifying these associations would provide another tool for water resource managers to use to help manage the threat to public health posed by cHABs and would be especially useful in the absence of microcystin-LR concentrations.  Thus, the goal of this paper is to utilize the National Lakes Assessment data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [@usepa2009national].  So that others may repeat or adjust this analysis, the data, code, and this manuscript are freely available via [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

@@

A detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for details.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  The 95% confidence intervals were calculated from 1000 bootstrapped samples. To identify chlorophyll *a* concentrations of concern we used a 50% conditional probability of exceeding each health advisory level and extracted the minimum chlorophyll *a* concentration that was associated with an upper confidence level of 50% or greater.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)`.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)` and the mean was `r mean(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)`.  Lastly, of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA childrens drinking water standard, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA adult drinking water standard,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO drinking water standard,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO recreational standard. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)` $\mu$g/L.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` and the mean was `r round(mean(nla_dat$CHLA,na.rm=T),2)`.  The associations between chlorophyll *a* and the upper confidence interval with a conditional probability of 50% were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]` for 0.3, 1.0, 1.6 and 2.0 $\mu$g/L, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

@@

There are numerous possible uses for the chlorophyll *a* and microcystin-LR advisory cut-off values.  First, in the absence of microcystin-LR measurements, exceedence of the chlorophyll *a* concentrations could be a trigger for further actions.  Given that there is uncertainity around these chlorophyll *a* cutoffs the best case scenario would be to monitor for chlorophyll and in the event of exceeding a target concentration take water samples and have those samples tested in a lab for microcystin-LR.

A second potential use is to identify possible bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanisms responsible for adverse impacts (e.g. toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs one approach has been to identify an increase over a baseline concentration of phycocyanin [@miller2013spatiotemporal].  This is a useful approach for targeted studies, but phycocyanin is also not always available and measures the predominance of cyanobacterial pigments and not toxins. Using our chlorophyll *a* cutoffs provides a value that is more directly associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.
